The perceptual dimensions of \\ sonority-driven epenthesis

### 2013-10-09 paper.tex
Second draft done

% faithfulness conflation?

In the final tally, the {\sc Sonority Angle}-based approach fares worse on one datapoint and uses approximately the same number of constraints, but manages to accomplish most of the work of the analysis with the only crucial ranking being the placement of a single {\sc Ident(Son$\measuredangle$)}$<$n constraint with respect to a relevant $\mathcal{M}$ constraint, while in Rose's analysis this work is spread over a number of different sonority-related constraints.

@@

For instance, it is well known that CV cues are much more robust than VC cues. %TODO : citation
This could

@@

\noindent {\bf Choice of horizontal distance.} In this paper, I assumed that the horizontal distance between C\textsubscript{1} and C\textsubscript{2}, and C\textsubscript{1} and V, is the same, and arbitrarily fixed at 1 unit, in calculating {\sc Sonority Angle}. This is an arbitrary choice, and potentially an important one, as when the horizontal distance gets large, changes occur in the cluster hierarchy, some of which violate the first desideratum for a sonority angle metric:

@@

\ex. Intensity data for English males (in dB) \citep[Table 5.13]{parker.2002}

@@

Using this scale, we obtain the desired empirical results of this paper based on the evidence of Irish and Chaha, that RN is the most epenthesisable of all the falling sonority clusters and tha RT and NT are the clusters most resistant to epenthesis. However, this does not conform to the general principle that for a given sonority distance, the more sonorous the consonants, the more likely a cluster is to undergo epenthesis --- we have NF being harder to epenthesise into than FT, here.

Moreover, this result is obtained by averaging stops and fricatives to ignore voicing. If we leave voicing intact, then these results disappear: fricative-voiceless stops are the most likely to undergo epenthesis of the falling sonority clusters, more so than liquid--nasal clusters, and if RT and NT fail to undergo epenthesis, we expect voiced fricative--voiced stop, nasal--voiceless fricative, and nasal--voiced stop clusters to also resist epenthesis. Both of these are contrary to the empirical results of this paper.

@@

\noindent{\bf Consonant deletion.} The current formulation of {\sc Ident(Son$\measuredangle$)}$<$n constraints restricts its application to when the consonants of a C\textsubscript{1}C\textsubscript{2} and C\textsubscript{1}VC\textsubscript{2} sequence stand in correspondence. However, another formulation is possible, in which sonority contours are a property of C\textsubscript{1} rather than a property of the whole contour.\footnote{Thanks to Adam Albright for this suggestion.}

@@

\noindent [This section remains to be written]

\ex. TODO: formulate this

A possible application of this looser definition of {\sc Sonority Angle} lies in the deletion of consonants in complex onsets. (pa vs pla) - loanwords, Sanskrit reduplication

>>>>

The more complex case of Chaha three- and four-consonant clusters also revealed that the {\sc Ident(Son$\measuredangle$)}$<$n constraints must be freely-rankable rather than fixed in a universal ranking. This allowed for faithfulness conflation of clusters both above and below a markedness constraint.

In the final tally, the {\sc Sonority Angle}-based approach fares worse on one datapoint and uses approximately the same number of constraints, but manages to accomplish most of the work of the analysis by ranking a single {\sc Ident(Son$\measuredangle$)}$<$n constraint with respect to a relevant $\mathcal{M}$ constraint, while in Rose's analysis this work is spread over a number of different sonority-related constraints.

@@

For instance, it is well known that CV cues are much more robust than VC cues \citep{wright.2004}.
This could be the reason for the offset of the consonant vastly outweighing the onset in terms of importance. Other sources for this asymmetry are also possible.

@@

\noindent {\bf Choice of horizontal distance.} In this paper, I assumed that the horizontal distance between C\textsubscript{1} and C\textsubscript{2}, and C\textsubscript{1} and V, is the same, and arbitrarily fixed at 1 unit, in calculating {\sc Sonority Angle}. This is an arbitrary choice, and potentially an important one. While the horizontal distance is of a similar magnitude to the values of the sonority scale, the {\sc Sonority Angles} are more or less as we expect, but when the horizontal distance is made very large, changes occur in the cluster hierarchy, some of which violate the first desideratum for a sonority contour distance metric. While it is still true that keeping C\textsubscript{1} fixed, raising C\textsubscript{2} decreases the {\sc Sonority Angle} (compare TR vs TN vs TF vs TT, for example), we no longer have a linear separation of the three broad classes of sonority rises, sonority plateaus and sonority falls.

@@

\ex. Intensity data for English males (in dB) \citep[Table 5.13]{parker.2002}\footnote{The intensity values measured for the three vowel classes are reversed from what one would expect according to phonological sonority \citep[214]{parker.2002}}, but overall this shows a good correlation with scales of phonological sonority. The distinction between some of these classes is not significant, including between the high and mid vowels.

@@

Using this scale, we obtain the desired empirical results of this paper based on the evidence of Irish and Chaha, that RN is the most epenthesisable of all the falling sonority clusters and that RT and NT are the clusters most resistant to epenthesis. However, this does not conform to the general principle that for a given sonority distance, the more sonorous the consonants, the more likely a cluster is to undergo epenthesis --- we have NF being harder to epenthesise into than FT, here.

Moreover, this result is obtained by averaging stops and fricatives to ignore voicing. If we leave voicing intact, then these results disappear: fricative--voiceless stops are the most likely to undergo epenthesis of the falling sonority clusters, more so than liquid--nasal clusters, and if RT and NT fail to undergo epenthesis, we expect voiced fricative--voiced stop, nasal--voiceless fricative, and nasal--voiced stop clusters to also resist epenthesis. Both of these are contrary to the empirical results of this paper.

@@

\noindent{\bf Consonant cluster simplification.} The current formulation of {\sc Ident(Son$\measuredangle$)}$<$n constraints restricts its application to when the consonants of a C\textsubscript{1}C\textsubscript{2} and C\textsubscript{1}VC\textsubscript{2} sequence stand in correspondence. However, another formulation is possible, in which offset sonority contours are a property of C\textsubscript{1} rather than a property of the whole contour.

\ex. \textsc{Ident(Son$\measuredangle$)}$<$n: For every input segment with an output correspondent, assign a violation mark if the angle between the offset sonority vectors from the segment in the input and output have a {\sc Sonority Angle} greater than $n$.

In revising this definition, we open up the possibility of applying {\sc Sonority Angle} to a larger class of phonological phenomena, such as consonant cluster simplification. For instance, in Sanskrit, reduplicants (underlined) with complex onsets are simplified:

\ex. \a. \underline{kan}-i-krand `cry out (intensive)'
     \c. \underline{sa}-swar `sound (perfective)'
     \c. \underline{tu}-stu `praise (perfective)'
     \z.
     \citep{steriade.1988} and \citep{parker.2002}

The observation is that the less sonorous of the two consonants is retained in the onset. What governs this choice? Traditionally, this has been analysed in terms of {\it the emergence of the unmarked}, with classes of consonants that are less marked in onset position --- i.e. less sonorous --- more likely to surface \citep{mccarthy.prince.1999}. However, another approach based on sonority contour faithfulness is also possible.

Consider the kan-krand case. The sonority contours to be considered are /kr/-[ka]. Both are rising sonority contours, so the {\sc Sonority Angle} will be fairly small at 0.12. Deletion of the /t/ in /stu/, on the other hand, would result in a falling sonority contour /st/ becoming a rising sonority contour [su], incurring a {\sc Sonority Angle} cost of 2.11. In both cases, deleting the first consonant would result in no violations of {\sc Ident(Son$\measuredangle$)}$<$n, according to the current definition, but this might violate constraints such as {\sc Left-Anchor}, which insist on an element on the left edge of a stem having a correspondent in the output \citep{mccarthy.prince.1999}. The following ranking thus predicts the deletion behaviour seen above.

\ex. {\sc Ident(Son$\measuredangle$)}$<$2.0 $\gg$ {\sc Left-Anchor}

\begin{center} \renewcommand*\arraystretch{1.2}
\scalebox{1}[1]{\begin{tabular}[t]{|rrl||c|c|} \hline 
\multicolumn{3}{|c||}{Input:~/\textipa{krand}/} & {\sc Ident(Son$\measuredangle$)}$<$2.0 & {\sc Left-Anchor} \\[0.5ex]
\hline \hline a. & \ding{43} & \textipa{kan} & 0.12 & \\
\hline b. & & \textipa{ran} & & $\ast$! \\
\hline \end{tabular}} \renewcommand*\arraystretch{1} \end{center}

\begin{center} \renewcommand*\arraystretch{1.2}
\scalebox{1}[1]{\begin{tabular}[t]{|rrl||c|c|} \hline 
\multicolumn{3}{|c||}{Input:~/\textipa{stu}/} & {\sc Ident(Son$\measuredangle$)}$<$2.0 & {\sc Left-Anchor} \\[0.5ex]
\hline \hline c. & \ding{43} & \textipa{su} & 2.11 $\ast$! & \\
\hline d. & & \textipa{tu} & & $\ast$ \\
\hline \end{tabular}} \renewcommand*\arraystretch{1} \end{center}

@@

To conclude this section, we see that the possible applications of sonority contour faithfulness are broad, encompassing many sonority-sensitive phonological processes. However, to pin down the exact parameters of the metric require more work.
